\section{Project Planning and Task Distribution}
\subsection{Work Breakdown}
To ensure traceability to the official grading rubric, the project was decomposed into major work packages that map directly to the required deliverables: FOL modeling, automatic knowledge-base generation, inference algorithms, baseline search algorithms, experimentation, and dissemination artifacts (report and video demonstration). Table~\ref{tab:work-breakdown} summarizes the work structure and expected outputs.

{\small
\begin{longtable}{|c|>{\RaggedRight\arraybackslash}p{4.8cm}|>{\RaggedRight\arraybackslash}p{8.6cm}|}
\caption{Project work breakdown aligned with assignment requirements}\label{tab:work-breakdown}\\
\hline
\textbf{WP} & \textbf{Major Task} & \textbf{Expected Deliverable} \\
\hline
\endfirsthead
\hline
\textbf{WP} & \textbf{Major Task} & \textbf{Expected Deliverable} \\
\hline
\endhead
\hline
\endfoot
\hline
\endlastfoot
WP1 & FOL formalization for Futoshiki & Complete set of axioms (domain, uniqueness, inequality, clue constraints), plus representative Skolemization and CNF derivations for the report. \\
\hline
WP2 & Automatic ground KB/CNF generation & Program module that automatically instantiates puzzle constraints for any grid size \(N\) and produces solver-ready clauses/facts. \\
\hline
WP3 & Forward Chaining solver & Rule-based inference engine implemented from scratch to derive valid value assignments and detect contradictions. \\
\hline
WP4 & Backward Chaining solver & Goal-driven inference mechanism (SLD-style querying) to infer \texttt{Val(i,j,v)} facts for unknown cells. \\
\hline
WP5 & A* search solver & Heuristic search implementation with admissible state evaluation and constraint-consistency pruning. \\
\hline
WP6 & Brute-force and Backtracking baselines & Two baseline solvers used for correctness checking and computational comparison. \\
\hline
WP7 & SAT-based solving support & SAT formulation and automated solving pipeline for cross-method verification and performance comparison. \\
\hline
WP8 & Experimental setup and I/O pipeline & Input/output standardization, benchmark script, and solved-instance export for all required test cases. \\
\hline
WP9 & Technical report and video demo & Final HCMUS-format report (analysis, experiments, discussion) and public demonstration video on representative puzzles. \\
\hline
\end{longtable}
}

\subsection{Member Responsibilities}
{\small
\begin{longtable}{|>{\RaggedRight\arraybackslash}p{4.1cm}|>{\RaggedRight\arraybackslash}p{10.1cm}|}
\caption{Member responsibilities}\label{tab:member-responsibilities}\\
\hline
\textbf{Member} & \textbf{Primary Responsibilities} \\
\hline
\endfirsthead
\hline
\textbf{Member} & \textbf{Primary Responsibilities} \\
\hline
\endhead
\hline
\endfoot
\hline
\endlastfoot
Nguyễn Khánh Toàn & A* search implementation; SAT-based solver integration; performance export/benchmark pipeline; repository coordination; README and video/demo coordination; GUI integration support; co-editing final report. \\
\hline
Nguyễn Chí Tài & Forward Chaining implementation; construction of input test files and puzzle-set preparation; support for report editing and validation tests. \\
\hline
Nguyễn Ngọc Thiên & FOL axiom design and CNF transformation workflow; Backward Chaining implementation; support for report editing and consistency checking. \\
\hline
Phan Quang Tiến & Automatic KB generation from CNF/grounding logic; Brute-force and Backtracking implementations; support for report editing and comparative analysis sections. \\
\hline
\end{longtable}
}

\subsection{Completion Percentage per Member}
For grading transparency, the team adopted a unified completion declaration at the end of development. As required by this submission, each member is recorded at \textbf{100\%} completion.

\begin{longtable}{|>{\RaggedRight\arraybackslash}p{8.8cm}|c|}
\caption{Completion percentage per member}\label{tab:completion-per-member}\\
\hline
\textbf{Member} & \textbf{Completion Percentage} \\
\hline
\endfirsthead
\hline
\textbf{Member} & \textbf{Completion Percentage} \\
\hline
\endhead
\hline
\endfoot
\hline
\endlastfoot
Nguyễn Khánh Toàn & 100\% \\
\hline
Nguyễn Chí Tài & 100\% \\
\hline
Nguyễn Ngọc Thiên & 100\% \\
\hline
Phan Quang Tiến & 100\% \\
\hline
\end{longtable}

\subsection{Project Management Board}
The team managed source code and integration through \textbf{GitHub} (branch-based collaboration, pull/merge workflow, and version traceability). Task planning and execution tracking were handled through a shared board (Trello/Planner-like workflow) with status labels (\textit{To do}, \textit{In progress}, \textit{Completed}), priority flags, assignees, and due dates. This process enabled transparent ownership, frequent synchronization, and early detection of integration risks.

Operationally, the management board was synchronized with technical milestones: formalization first, then solver development, then benchmarking and reporting. This ordering reduced late-stage rework and ensured that rubric-weighted items were completed before final packaging.

\clearpage

{\small
\begin{longtable}{|>{\RaggedRight\arraybackslash}p{2.3cm}|>{\RaggedRight\arraybackslash}p{2.6cm}|>{\RaggedRight\arraybackslash}p{3.5cm}|>{\RaggedRight\arraybackslash}p{4.5cm}|>{\RaggedRight\arraybackslash}p{1.8cm}|}
\caption{Project timeline and progress board summary}\label{tab:timeline-board}\\
\hline
\textbf{Period} & \textbf{Stage} & \textbf{Main Activities} & \textbf{Outputs} & \textbf{Status} \\
\hline
\endfirsthead
\hline
\textbf{Period} & \textbf{Stage} & \textbf{Main Activities} & \textbf{Outputs} & \textbf{Status} \\
\hline
\endhead
\hline
\endfoot
\hline
\endlastfoot
08/04/2026--\newline 11/04/2026 & Planning and formalization & Requirement analysis, role assignment, FOL specification, initial CNF workflow design. & Work plan, formal FOL draft, initial puzzle inputs. & Completed \\
\hline
12/04/2026--\newline 17/04/2026 & Core algorithm implementation & Forward/Backward Chaining, A*, SAT pipeline, KB generation, backtracking baseline modules. & Implemented solver modules and intermediate validation runs. & Completed \\
\hline
18/04/2026--\newline 20/04/2026 & Integration and benchmark & Unified I/O handling, solved-output export, runtime comparison, GUI connection and stability checks. & End-to-end executable system and benchmark artifacts. & Completed \\
\hline
21/04/2026--\newline 25/04/2026 & Documentation and delivery & Report refinement, figure/table completion, demo video finalization, submission packaging. & Final report PDF, demo link, and complete submission package. & Completed \\
\hline
\end{longtable}
}

