\subsection{Backward Chaining (SLD Resolution)}

% Approach & Architecture
\subsubsection{Idea}
The implemented solver is a hybrid architecture that combines Prolog-style SLD resolution with explicit state-space backtracking.
\begin{itemize}
    \item \textbf{KB construction ($HornConverter$):} Converts a puzzle instance into Horn facts and rules using predicates such as $val(i,j,v)$, $given(i,j,v)$, and $not\_val(i,j,v)$.
    \item \textbf{Backward chaining core ($fol\_bc\_or$ / $fol\_bc\_and$):} A Prolog-like SLD interpreter with unification, substitution, DFS goal reduction, cycle detection, and NAF handling.
    \item \textbf{Search controller ($\_backtrack$):} Controls assignment of values to cells. At each step, it selects one unassigned cell, queries $val(r,c,v)$ to obtain legal candidates, asserts one candidate as a temporary fact, recurses, and retracts on failure.
\end{itemize}

\paragraph{Why search is integrated with SLD Resolution:}
The axiom $\forall i\forall j\exists v\;Val(i,j,v)$ contains existential quantification and cannot be encoded directly as a definite Horn rule. Therefore, the solver uses backtracking for value choice, while the backward chaining engine is used to validate and prune candidates through derivable constraints (mainly $not\_val$).

\paragraph{Technical Implementation Notes:}~
\begin{itemize}
    \item \textbf{0-based indexing in code:} Internally, rows/columns are represented as $0..N-1$ for direct mapping to Python lists.
    \item \textbf{Cell-level candidate generation with NAF:} The generator rule $val(i,j,v) \leftarrow domain(v),\; naf(not\_val(i,j,v))$ is used to propose candidates that are not provably forbidden.
    \item \textbf{Hybrid workflow:} BC serves as a logical inference engine; backtracking constructs a full solution by incrementally asserting/retracting $val$ facts.
\end{itemize}

\subsubsection{Algorithm Analysis}
% Knowledge Representation
\paragraph{Atoms (Predicates)}~\\
The Horn KB used the following vocabulary:
\begin{itemize}
    \item $val(i,j,v)$: Cell $(i,j)$ is assigned the value $v$.
    \item $given(i,j,v)$: Cell $(i,j)$ has the pre-filled clue value $v$
    \item $less\_h(i,j)$ / $greater\_h(i,j)$: Horizontal inequalities between $(i,j)$ and $(i,j+1)$.
    \item $less\_v(i,j)$ / $greater\_v(i,j)$: Vertical inequalities between $(i,j)$ and $(i+1,j)$.
    \item $less(v1,v2)$: Background arithmetic fact representing $v_1 < v_2$.
    \item $neq(x,y)$: Background inequality fact representing $x \neq y$.
    \item $not\_val(i,j,v)$: Value $v$ is explicitly forbidden at cell $(i,j)$.
\end{itemize}

\paragraph{Facts}~
\begin{itemize}
    \item \textbf{Background facts}: All valid $less(v1,v2)$ and $neq(x,y)$ combinations within the finite domain $1..N$.
    \item \textbf{Instance facts}: $given(...)$ clues, inequality signs, and immediate "boundary pruning" facts ($not\_val(...)$) derived directly from extreme inequality conditions (e.g., a cell on the smaller side of an inequality cannot hold the maximum value $N$).
\end{itemize}

\paragraph{Rules (Definite Horn)}~\\
Constraints are encoded by Horn rules that derive either $val$ (from givens) or $not\_val$ (forbidden assignments):
\begin{itemize}
    \item Given $\Rightarrow$ $Val: val(i,j,v) :- given(i,j,v)$
    \item Cell uniqueness: If a value is chosen, all others are forbidden in that cell.
    \item Row/Col uniqueness: If $val(i,j1,v)$ holds, then $not_val(i,j2,v)$ for all $j_2 \neq j_1$.
    \item Inequalities: Values that violate a < or > constraint trigger a $not\_val$ derivation.
\end{itemize}

\paragraph{Existence / Choice Rule via NAF}~\\
To generate candidates for an empty cell, the solver adds:
\[
val(i,j,v) \leftarrow domain(v),\; naf(not\_val(i,j,v)).
\]
A value $v$ is accepted as a candidate for $(i,j)$ when $v$ is in the domain and $not\_val(i,j,v)$ cannot be proven.



% Core Mechanisms
\paragraph{Unification and Substitution}~
\begin{itemize}
    \item $unify(x,y,\theta)$: Computes the Most General Unifier (MGU) for two expressions, returning the substitution dictionary $\theta$ or "failure".
    \item $subst(\theta, expr)$: Applies the substitution $\theta$ to goals and rule bodies. Variables are represented as lowercase strings (e.g., 'v', 'i', 'j').
\end{itemize}

\paragraph{SLD Resolution (AND/OR Search Tree)}~
\begin{itemize}
    \item OR node ($bc\_or$): Proves a goal by attempting to match and unify it against the head of any available fact or rule.
    \item AND node ($bc\_and$): Proves the body of a rule by proving all subgoals sequentially from left to right.
    \item Evaluation follows depth-first, leftmost goal expansion.
\end{itemize}

\paragraph{Negation as Failure (NAF) Handling}~
\begin{itemize}
    \item $naf(G)$ succeeds iff the engine fails to derive $G$.
    \item Inside NAF context ($in\_naf=True$), rule expansion for predicate $val$ is disabled. This prevents self-supporting loops between $val$ and $naf(not\_val)$.
\end{itemize}


% Optimizations
\paragraph{Optimizations Implemented}~
\begin{itemize}
    \item \textbf{MRV variable ordering:} Backtracking chooses the unassigned cell with minimum remaining values (MRV) based on current BC query results.
    \item \textbf{Predicate indexing:} Facts and rules are indexed by predicate ($fact\_index$, $pred\_rule\_index$, $exact\_rule\_index$).
    \item \textbf{Cell-specific index for $val$:} Additional indexes by cell $(i,j)$ are used for both facts and rules of predicate $val$, reducing scan cost for queries of the form $val(r,c,v)$.
    \item \textbf{Versioned memoization:} Memo keys include $kb\_version$, so cache is reused across queries while remaining correct after assert/retract operations.
    \item Cycle Detection (visited set): Tracks the active resolution path to prevent infinite recursion in cyclic goal graphs.
    \item Boundary Pruning: Converting extreme inequalities into immediate $not\_val$ facts drastically reduces the search space before resolution even begins.
\end{itemize}


\subsubsection{Pseudocode}
% ham solve
\begin{algorithm}[H]
\caption{Master Coordinator}
\label{alg:label}
\begin{algorithmic}
\Function {solve}{\textit{puzzle}}
\State $kb \gets HornConverter(puzzle).build()$
\State add facts $domain(1..N)$
\State add facts $val(r,c,v)$ for all givens
\State add generator rules $val(i,j,v) :- domain(v), naf(not\_val(i,j,v))$
\State build fact/rule indexes and cell indexes for $val$
\State initialize $kb\_version \gets 1$

\For{each given $(r, c) = v$}
\If {$FOL\_BC\_ASK(kb, not\_val(r, c, v))$ succeeds}
\State\Return failure // Contradiction in givens
\EndIf
\EndFor
\State\Return $\_BACKTRACK(kb)$
\EndFunction
\end{algorithmic}
\end{algorithm}

% ham backtrack
\begin{algorithm}[H]
\caption{State Space Manager}
\begin{algorithmic}[1]
\Function{Backtrack}{$kb$}
    \If{grid complete}
        \State \Return $success$
    \EndIf
    \State $(r,c,\mathcal{C}) \gets$ MRV-SelectCell($kb$)
    \If{$\mathcal{C} = \emptyset$}
        \State \Return $failure$
    \EndIf
    \For{each $value$ in $\mathcal{C}$}
        \State $grid[r][c] \gets value$
        \State assert fact $val(r,c,value)$ into $kb$
        \State $kb\_version \gets kb\_version + 1$
        
        \If{\Call{\_Backtrack}{$kb$} $success$}
            \State \Return $success$
        \EndIf
        
        \State retract fact $val(r,c,value)$ from $kb$
        \State $kb\_version \gets kb\_version + 1$
        \State $grid[r][c] \gets 0$
    \EndFor
    \State \Return $failure$
\EndFunction
\end{algorithmic}
\end{algorithm}

% hĂ m fol_bc_ask 
\begin{algorithm}[H]
\caption{Inference Interface}
\begin{algorithmic}[1]
\Function{fol\_bc\_ask}{$kb, goal$}
    \State \Return \Call{fol\_bc\_or}{$kb, goal, theta=\emptyset, in\_naf=false, visited=\emptyset$}
\EndFunction
\end{algorithmic}
\end{algorithm}

% hĂ m fol_bc_or
\begin{algorithm}[H]
\caption{Goal Resolver - OR Node}
\begin{algorithmic}[1]
\Function{fol\_bc\_or}{$kb, goal, theta, in\_naf, visited$}
    \State $g \gets \Call{Subst}{\theta, goal}$
    \If{$(g, in\_naf) \in visited$}
        \State \Return $\emptyset$ \Comment{Cycle detection}
    \EndIf
    \State $visited' \gets visited \cup \{(g, in\_naf)\}$
    \State $k \gets (kb\_version, g, in\_naf, \theta)$
    \If{$k \in memo$}
        \State \Return $memo[k]$
    \EndIf

    \If{$g$ is $naf(inner)$} \Comment{Negation As Failure}
        \If{\Call{fol\_bc\_or}{$kb, inner, theta, true, visited'$} yields no proof}
            \State \Return $\{theta\}$ \Comment{NAF success}
        \Else
            \State \Return $\emptyset$ \Comment{NAF failure}
        \EndIf
    \EndIf

    \State $answers \gets \emptyset$
    \State $facts \gets$ indexed facts matching $pred(g)$
    \If{$pred(g)=val$ and $(i,j)$ of $g$ is grounded}
        \State $facts \gets$ cell-indexed facts for $(i,j)$
    \EndIf
    \For{each fact $F$ in $facts$}
        \State $theta' \gets \Call{Unify}{args(g), args(head(F)), theta}$
        \If{$theta' \neq failure$}
            \State $answers \gets answers \cup \Call{fol\_bc\_and}{kb, body(F), theta', in\_naf, visited'}$
        \EndIf
    \EndFor

    \If{\textbf{not} ($in\_naf$ and $pred(g)=val$)}
        \State choose rule candidates from exact/predicate/cell index
        \For{each rule $R$ in candidates}
            \State $theta' \gets \Call{Unify}{args(g), args(head(R)), theta}$
            \If{$theta' \neq failure$}
                \State $answers$ $\gets$ $answers \cup \Call{fol\_bc\_and}{kb, body(R), theta', in\_naf, visited'}$
            \EndIf
        \EndFor
    \EndIf

    \State $memo[k] \gets answers$
    \State \Return $answers$
\EndFunction
\end{algorithmic}
\end{algorithm}

% hĂ m fol_bc_and
\begin{algorithm}[H]
\caption{Rule Body Validator - AND Node}
\begin{algorithmic}[1]
\Function{fol\_bc\_and}{$kb, goals, theta, in\_naf, visited$}
    \If{$goals$ is empty}
        \State \Return $\{theta\}$
    \EndIf
    \State $first \gets \Call{Subst}{theta, goals[0]}$
    \State $rest \gets goals[1:]$
    \State $answers \gets \emptyset$
    
    \For{each $theta'$ in \Call{fol\_bc\_or}{$kb, first, theta, in\_naf, visited$}}
        \State $answers \gets answers \cup \Call{fol\_bc\_and}{kb, rest, theta', in\_naf, visited}$
    \EndFor
    
    \State \Return $answers$
\EndFunction
\end{algorithmic}
\end{algorithm}

\subsubsection{Illustration}
On a query $val(r,c,v)$ for an unassigned cell, the resolver first tries indexed $val$ facts for $(r,c)$, then indexed $val$ rules for $(r,c)$ (plus generic generator rules). The result is a candidate set $\mathcal{C}$.

MRV chooses the cell with smallest $|\mathcal{C}|$. If a candidate $v \in \mathcal{C}$ is tentatively assigned, the solver asserts $val(r,c,v)$ into KB and increments $kb\_version$. This makes old memo entries automatically inactive (because memo key includes $kb\_version$), while still preserving cache reuse for repeated subgoals inside the same KB state.

If recursion fails, the fact is retracted, $kb\_version$ is incremented again, and the next candidate is tried. This loop continues until a complete grid is found or the branch set is exhausted.

