\subsection{SAT Solver (Using SAT Library)}
\subsubsection{Idea}
The SAT approach converts Futoshiki constraints into CNF clauses and delegates solving to PySAT (Glucose3). In this project, \texttt{FOLKB} builds the CNF, then the SAT model is decoded back to an \(N \times N\) grid.

\subsubsection{Constraint Encoding}
To avoid repeating full FOL derivations from Section 3.2, we only summarize the SAT-level encoding. For each \((i,j,v)\), define:
\[
X_{i,j,v} = \text{true} \iff \text{cell }(i,j)\text{ has value }v.
\]

Variable ID mapping used in code:
\[
\text{ID}(i,j,v)= i\cdot N^2 + j\cdot N + v,
\]
with \(i,j\in[0,N-1]\), \(v\in[1,N]\), so IDs are in \([1,N^3]\).

Main CNF groups are:
\begin{itemize}
	\item each cell has exactly one value;
	\item row and column uniqueness (Latin-square constraints);
	\item given clues as unit clauses;
	\item inequality constraints by forbidding violating adjacent value pairs.
\end{itemize}
Thus, any SAT model corresponds to a valid solution.

\subsubsection{Pseudocode}
\begin{algorithm}[H]
\caption{SAT-based Futoshiki Solver}
\begin{algorithmic}[1]
\State Build CNF clauses from puzzle constraints using FOLKB
\State Create SAT solver instance (Glucose3)
\State Add all CNF clauses to solver
\If{solver returns UNSAT}
	\State \Return \textbf{failure}
\EndIf
\State Get SAT model (set of literals)
\State Initialize empty grid \(N \times N\)
\For{each positive literal \(\ell\) in model}
	\If{\(1 \le \ell \le N^3\)}
		\State Decode \(\ell \rightarrow (i,j,v)\)
		\State Set grid\((i,j) \leftarrow v\)
	\EndIf
\EndFor
\If{any cell remains unassigned}
	\State \Return \textbf{failure}
\Else
	\State \Return solved grid
\EndIf
\end{algorithmic}
\end{algorithm}

\subsubsection{Illustration}
For a \(4 \times 4\) puzzle:
\begin{enumerate}
	\item Create \(4^3=64\) variables \(X_{i,j,v}\).
	\item Add clauses for uniqueness, clues, and inequalities.
	\item Run Glucose3 and decode the satisfying literals.
	\item Reconstruct the solved grid.
\end{enumerate}

\subsubsection{Algorithm Analysis}
\textbf{Encoding cost:}
\begin{itemize}
	\item Number of variables is \(N^3\).
	\item Number of clauses is polynomial in \(N\), mainly from exclusivity and inequality constraints.
\end{itemize}

\textbf{Solving cost:}
\begin{itemize}
	\item SAT is NP-complete in general, so worst-case complexity is exponential.
	\item In practice, CDCL solvers are effective on structured instances.
\end{itemize}

\textbf{Space and implementation trade-offs:}
\begin{itemize}
	\item Requires storing full CNF and SAT internal structures.
	\item Easy to extend by adding new clause templates.
\end{itemize}

\textbf{Practical behavior in this project:}
\begin{itemize}
	\item Provides clear SAT/UNSAT feasibility checking.
	\item Runtime is competitive in experiments, but SAT is not the central focus of this report.
\end{itemize}
